\chapter{Compatibility with CMU \smv}
The \nusmv language is mostly source compatible with the original
version of \smv distributed at Carnegie Mellon University from
which we started.  In this appendix we describe the most common
problems that can be encountered when trying to use old CMU
\smv programs with \nusmv.

The main problem is variable names in old programs that conflicts with
new reserved words.  The list of the new reserved words of \nusmv
w.r.t. CMU \smv is the following:

\begin{table}[h]
\begin{tabular}{p{100pt}p{300pt}}
\code{F, G, X, U, V, W, H, O, Y, Z, S, T, B} &
These names are reserved for the LTL temporal operators.\\
\code{LTLSPEC} &
It is used to introduce LTL specifications. \\
\code{INVARSPEC} &
It is used to introduce invariant specifications.\\
\code{IVAR} &
It is used to introduce input variables. \\
\code{JUSTICE} &
It is used to introduce ``justice'' fairness constraints.\\
\code{COMPASSION} &
It is used to introduce ``compassion'' fairness constraints. \\
\end{tabular}
\end{table}

The \code{IMPLEMENTS}, \code{INPUT}, \code{OUTPUT} statements are not
supported by \nusmv. They are parsed from the input file, but 
are internally ignored.

\nusmv differs from CMU \smv also in the controls that are
performed on the input formulas. Several formulas that are valid for CMU
\smv, but that have no clear semantics, are not accepted by 
\nusmv. In particular:

\begin{itemize}
\item It is no longer possible to write formulas containing nested
`\code{next}'.
\begin{alltt}
TRANS
  next(alpha & next(beta | next(gamma))) -> delta
\end{alltt}
\item It is no longer possible to write formulas containing
`\code{next}' in the right hand side of ``normal'' and ``init''
assignments (they are allowed in the right hand side of ``next''
assignments), and with the statements `\code{INVAR}' and
`\code{INIT}'.
\begin{nusmvCode}
INVAR
  next(alpha) & beta
INIT
  next(beta) -> alpha
ASSIGN
  delta := alpha & next(gamma);       -- normal assignments
  init(gamma) := alpha & next(delta); -- init assignments
\end{nusmvCode}
\item It is no longer possible to write `\code{SPEC}',
`\code{FAIRNESS}' statements containing `\code{next}'.
\begin{nusmvCode}
FAIRNESS
 next(running)
SPEC
 next(x) & y
\end{nusmvCode}
\item The check for circular dependencies among variables has been 
done more restrictive.
We say that variable {\it x} depends on variable {\it y} if {\it x := f(y)}.
We say that there is a circular dependency in the definition of {\it x} if:
\begin{itemize}
\item  {\it x} depends on itself ( e.g. {\it x := f(x,y)} );
\item {\it x} depends on {\it y} and {\it y} depends on {\it x} (e.g. {\it x :=
f(y)} and {\it y := f(x)} or {\it x := f(z)}, {\it z := f(y)} and {\it y :=
f(x)} ).
\end{itemize}
In the case of circular dependencies among variables there is no fixed
order in which we can compute the involved variables. Avoiding circular
dependencies among variables guarantee that there exists an order in
which the variables can be computed. In \nusmv circular
dependencies are not allowed.

In CMU \smv the test for circular dependencies is able to detect
circular dependencies only in ``normal'' assignments, and not in ``next''
assignments. The circular dependencies check of \nusmv has been
extended to detect circularities also in ``next'' assignments. For
instance the following fragment of code is accepted by CMU \smv
but discarded by \nusmv.
\begin{nusmvCode}
MODULE main
VAR
  y : boolean;
  x : boolean;
ASSIGN
  next(x) := x & next(y);
  next(y) := y & next(x);
\end{nusmvCode}
\end{itemize}

Another difference between \nusmv and CMU \smv is in the variable
order file.  The variable ordering file accepted by \nusmv can be
partial and can contain variables not declared in the model. Variables
listed in the ordering file but not declared in the model are simply
discarded. The variables declared in the model but not listed in the
variable file provided in input are created at the end of the given
ordering following the default ordering. All the ordering files
generated by CMU \smv are accepted in input from \nusmv but the
ordering files generated by \nusmv may be not accepted by CMU \smv.
Notice that there is no guarantee that a good ordering for CMU \smv is
also a good ordering for \nusmv.  In the ordering files for \nusmv,
identifier \code{\_process\_selector\_} can be used to control the
position of the variable that encodes process selection. In CMU \smv
it is not possible to control the position of this variable in the
ordering; it is hard-coded at the top of the ordering.
